## Reproducible Research Course Project 1
Submitted by: **Josh Valdeleon**

Date Committed: December 4, 2020


### Abstract
In this project, we will be working with the data on personal movements, at the very least, we'll work with the number of steps recorded with **5-minute** interval from **October** to **November 2012**. 

The data is stored in a comma-separated-value (CSV) file which can be downloaded in this [link](https://github.com/imjbmkz/RepData_PeerAssessment1/blob/master/activity.zip). It consists of **17,568** observations and **3** variables. These variables are as shown.

|Variable Name                  |Description                            |
|-------------------------------|---------------------------------------|
|`steps`                        |Number of steps taking in a 5-minute interval (missing values are coded as **NA**
|`date`                         |The date on which the measurement was taken in YYYY-MM-DD format
|`interval`                     |Identifier for the 5-minute interval in which measurement was taken

The analysis will revolve around answering couple of questions and presenting some data visualizations which would help in understanding further the data.

The only libraries used are **readr** to efficiently read large data, **dplyr** for data manipulation, and **ggplot2** for data visualization. These packages are all included in the **tidyverse** package.

All of the codes are stored in the R script called `project1_solutions.R`. This script provides five functions which would help answering the questions presented. I have sourced this script and used these functions to answer the problems below. For more details, go to `README.md` is this sub-directory.


```{r cache=TRUE, results='hide'}
source('~/reproducible_research/project1_solutions.R')
```


### Loading and Pre-Processing the data
Let's read the data and store it in variable `x`.

```{r cache=TRUE, results='hide'}
x <- get_data()
```


Let's quickly examine the data.

```{r cache=TRUE}
print(names(x)) # gives the column names
print(dim(x)) # gives the number of observations and variables
print(str(x)) # gives the 'structure' of the data
```


By the above functions, we've seen the data types of these columns, its size, and some `NA`s. We'll not touch these for now as it will be addressed during question number three.

### Question 1: What is mean total number of steps taken per day?

```{r cache=TRUE}
q1 <- question_one(x)
```


The function above is still part of the script that I sourced. `q1` is a list with 4 named-attributes. Let's call the ggplot object from this variable. It's called `my_hist`.

```{r cache=TRUE}
q1$my_hist
```


The histogram shows us the distribution of the values in `steps` variable of our dataset. This is after removing all `NA`s from it. It looks close to a normal distribution which can be proved by the mean and median of this dataset. 

```{r cache=TRUE}
print(paste('The mean is', round(q1$my_mean, 2)))
print(paste('The median is', q1$my_median))
```


The mean and median are so close to each other. The mean and median of a data that has a **normal distribution** are equal. Although ours aren't but they're too close. 


## Question 2: What is the average daily activity pattern?

```{r cache=TRUE}
q2 <- question_two(x)
```


The time-series plot would helps us easily see which interval has the highest mean steps across all days. It's stored in the named item in `q2` called `my_line`.

```{r cache=TRUE}
q2$my_line
```


The peak of the line is on roughly **800th minute**. And to be more precise, we simple call the second element of `q2` list, that is, `my_max`.

```{r cache=TRUE}
q2$my_max
```


Our estimate is correct, and the exact interval where we've seen the highest mean steps across all intervals is on the **835<sup>th</sup>** interval.


## Imputing missing values

```{r cache=TRUE}
q3 <- question_three(x)
```


Let's look at the rate of missing values per column of the original dataset.

```{r cache=TRUE}
q3$na_prop
```


The `steps` variable is the only column with missing values that is more than 13% of the data. To address this, I tried to do two approaches. One is to replace the `NA`s with the **mean** of this variable, and two is to replace the same with the **median**. 

The function `question_three()` mutates the data with two additional columns with `NA` values replaced by mean and median. They are called `mean_filled` and `median_filled`. Let's look at the mean and median of these two columns.

```{r cache=TRUE}
print('mean_filled')
print(str(q3$mean_filled))
print('median_filled')
print(str(q3$median_filled))
```


From above output, we can see that the mean and median will become the same if the `NA`s are replaced with the mean of the `steps` variable. On the other hand, the mean becomes smaller than the median if these `NA`s were replaced by the median. To visualize it, `q3` list has ggplot objects called `mean_filled_hist` and `median_filled_hist` which will display the histograms of `mean_filled` and `median_filled` variables.

```{r cache=TRUE}
gridExtra::grid.arrange(q3$mean_filled_hist, q3$median_filled_hist)
```


We can see that this histogram proves the calculated mean and median of `mean_filled` since it looks like (and numerically speaking, it is) a **normal distribution**. The `median_filled` variable looks like a **bimodal** one. In many statistical analyses, a normal distribution would be much easier to model than other distributions. So, in the next question, we'll replace the `NA`s with the mean steps.


## Are there differences in activity patterns between weekdays and weekends?

```{r echo=TRUE}
q4 <- question_four(x)
```


`question_four()` function replaces `NA`s in the `steps` variable, and adds a new column that categorizes the date if it is a weekday or a weekend. The data is then grouped by this category and by interval, and summarized by the mean steps.

Let's visualize the trend of the mean steps per interval, and we'll group it by day category.

```{r echo=TRUE}
q4$my_line
```


From the above time-series plot, I would say that generally, the trend is similar. The person is much more active, that is, taking more steps roughly at **800<sup>th</sup>** minute. The difference between them is that the person is much more active during weekends since there are many intervals after 1000<sup>th</sup> were s/he exceeded 100 steps versus that of the weekdays where s/he only had one instance of reaching 100 steps. 

This could mean that maybe the person during weekdays, which is obviously his/her workdays, might always be at his/her desk and just walks when needed (for lunch, going to different meeting rooms, goes to his/her officemates to have a little chit-chat, etc). While on the weekend, that his/her time to go on a trip or elsewhere to have fun.